---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# tidyLPA

The goal of `tidyLPA` is to provide tools to make it easier to use the `R` package `MCLUST` for Latent Profile Analysis analyses.

This is a sister-project to [prcr](https://github.com/jrosen48/prcr), for two-step cluster analysis. 

tidyLPA has been benchmarked to LPA, at least for a simple dataset (the [iris dataset](https://en.wikipedia.org/wiki/Iris_flower_data_set)). You can find the results of the benchmarking [here](https://jrosen48.github.io/blog/comparing-mplus-and-mclust-output/). Further benchmarking with data that are more complex and particularly data that are not as normally distributed as the continuous variables in the iris data set are is an important issue and future direction.

# Example

First, we can explore the Bayesian Information Criteria (BIC) or the Integrated Complete-data Likelihood (ICL) values, using the `explore_models_clust()` function:

```{r, echo = F}
devtools::load_all(".")
```

```{r, eval = F}
library(tidyLPA)
```

Using the built-in `pisaUSA15` dataset and variables for broad interest, instrumental motivation, and self-efficacy, we can quickly explore a three profile solution:

```{r}
d <- pisaUSA15
m3 <- create_profiles_mclust(d, broad_interest, instrumental_mot, self_efficacy, n_profiles = 3, to_return = "tibble")
summary_mclust(m3)
plot_profiles_mclust(m3)
```

We can also explore a range of models:

```{r}
explore_models_mclust(d, broad_interest, instrumental_mot, self_efficacy)
```
