---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-"
)
```

# tidyLPA

[![Travis build status](https://travis-ci.org/jrosen48/tidyLPA.svg?branch=master)](https://travis-ci.org/jrosen48/tidyLPA)

tidyLPA provides the functionality to carry out Latent Profile Analysis. Note that tidyLPA is still at the beta stage! Please report any bugs at https://github.com/jrosen48/tidyLPA or send an email to [jrosen@msu.edu](mailto:jrosen@msu.edu).

## Installation

You can install tidyLPA from GitHub with:

```{r gh-installation, eval = FALSE}
install.packages("devtools")
devtools::install_github("jrosen48/tidyLPA")
```

## Example

Here is a brief example using the built-in `pisaUSA15` dataset and variables for broad interest, enjoyment, and self-efficacy. See `?create_profiles_lpa` for more details.

```{r}
library(tidyLPA)
```

```{r, warning = F}
d <- pisaUSA15[1:100, ]

m3 <- estimate_profiles_lpa(d, broad_interest, enjoyment, self_efficacy, n_profiles = 3, model = 2)

plot_profiles_lpa(m3, to_center = TRUE)
```

# Learn more

Please check out the [tidyLPA website](https://jrosen48.github.io/tidyLPA/) to learn more.

In particular, the [Intro to tidyLPA vignette](https://jrosen48.github.io/tidyLPA/articles/Introduction_to_tidyLPA.html) has more in-depth information.

Please note that this project is released with a [Contributor Code of Conduct](CONDUCT.md). By participating in this project you agree to abide by its terms.

# Contact

If you have any questions or feedback, please do not hesitate to get in touch:

* By [email](mailto:jrosen@msu.edu)
* By [Twitter](http://twitter.com/jrosenberg6432)
* Through filing an issue on GitHub [here](https://github.com/jrosen48/tidyLPA)
