---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-",
  warning = T
)
```

# tidyLPA

The goal of `tidyLPA` is to provide tools to make it easier to use the `R` package [MCLUST](http://www.stat.washington.edu/mclust/) for Latent Profile Analysis analyses.

This is a sister-project to [prcr](https://github.com/jrosen48/prcr), for two-step cluster analysis. 

tidyLPA (which, again, is an interface to the `MCLUST` package) has been benchmarked to MPlus, at least for a simple dataset (the [iris dataset](https://en.wikipedia.org/wiki/Iris_flower_data_set)). You can find the results of that benchmarking, which showed the results to be nearly, identical, [here](https://jrosen48.github.io/blog/comparing-mplus-and-mclust-output/). 

# Example

First, we can explore the Bayesian Information Criteria (BIC) or the Integrated Complete-data Likelihood (ICL) values, using the `explore_models_clust()` function:

```{r, echo = F}
devtools::load_all(".")
```

```{r, eval = F}
library(tidyLPA)
```

Using the built-in `pisaUSA15` dataset and variables for broad interest, instrumental motivation, and self-efficacy, we can quickly explore a three profile solution:

```{r}
d <- pisaUSA15
m3 <- create_profiles_mclust(d, broad_interest, instrumental_mot, self_efficacy, n_profiles = 3, to_return = "tibble")
summary_mclust(m3)
plot_profiles_mclust(m3)
```

We can also explore a range of models:

```{r}
explore_models_mclust(d, broad_interest, instrumental_mot, self_efficacy)
```
