---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# tidyLPA

The goal of `tidyLPA` is to provide tools to make it easier to use the `R` package [MCLUST](http://www.stat.washington.edu/mclust/) for Latent Profile Analysis analyses.

## Installation

You can install tidyLPA from GitHub with:

```{r gh-installation, eval = FALSE}
# install.packages("devtools")
devtools::install_github("jrosen48/tidyLPA")
```

# Example

Here is a brief example using the built-in `pisaUSA15` dataset and variables for broad interest, enjoyment, and self-efficacy. See `?create_profiles_lpa` for more details.

Also, please view the [Introduction to tidyLPA vignette](https://jrosen48.github.io/tidyLPA/articles/Introduction_to_tidyLPA.html) for more information.

```{r}
library(tidyLPA)
```

```{r, warning = F}
library(dplyr, warn.conflicts = FALSE)
d <- pisaUSA15[1:100, ]
m3 <- create_profiles_lpa(d, broad_interest, enjoyment, self_efficacy, n_profiles = 3, model = 2)
plot_profiles_lpa(m3)
```
