---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-",
  warning = T, 
  error = F
)
```

# tidyLPA

The goal of `tidyLPA` is to provide tools to make it easier to use the `R` package [MCLUST](http://www.stat.washington.edu/mclust/) for Latent Profile Analysis analyses.

This is a sister-project to [prcr](https://github.com/jrosen48/prcr), for two-step cluster analysis. 
tidyLPA (which, again, is an interface to the `MCLUST` package) has been benchmarked to MPlus, at least for a simple dataset (the [iris dataset](https://en.wikipedia.org/wiki/Iris_flower_data_set)). You can find the results of that benchmarking, which showed the results to be nearly, identical, [here](https://jrosen48.github.io/blog/comparing-mplus-and-mclust-output/). 

# Example

Here is a brief example using the built-in `pisaUSA15` dataset and variables for broad interest, enjoyment, and self-efficacy. See `?create_profiles_lpa` for more details.

Also, please view the [Introduction to tidyLPA vignette](https://jrosen48.github.io/tidyLPA/articles/Introduction_to_tidyLPA.html) for more information.

```{r}
library(tidyLPA)
```

```{r, warning = F}
library(dplyr)
d <- pisaUSA15
d <- sample_frac(d, .5)
m3 <- create_profiles_lpa(d, broad_interest, enjoyment, self_efficacy, n_profiles = 3, model = 2)
m3
plot_profiles_lpa(m3, to_center = TRUE)
```
