---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-"
)
```

# tidyLPA

[![Travis build status](https://travis-ci.org/jrosen48/tidyLPA.svg?branch=master)](https://travis-ci.org/jrosen48/tidyLPA) [![CRAN status](http://www.r-pkg.org/badges/version/tidyLPA)](https://cran.r-project.org/package=tidyLPA)

tidyLPA provides the functionality to carry out Latent Profile Analysis (LPA). LPA is a statistical modeling approach for estimating parameters (i.e., the means, variances, and covariances) distinct profiles of groups of variables. Note that tidyLPA is still at the beta stage! Please report any bugs at https://github.com/jrosen48/tidyLPA or send an email to [jrosen@msu.edu](mailto:jrosen@msu.edu).

## Installation

You can install tidyLPA from CRAN with:

```{r, eval = F}
install.packages("tidyLPA")
```

You can also install the in-development version of tidyLPA from GitHub with:

```{r gh-installation, eval = FALSE}
install.packages("devtools")
devtools::install_github("jrosen48/tidyLPA")
```

## Example

Here is a brief example using the built-in `pisaUSA15` dataset and variables for broad interest, enjoyment, and self-efficacy. See `?estimate_profiles` for more details.

```{r}
library(tidyLPA)
```

```{r}
d <- pisaUSA15[1:100, ]

m3 <- estimate_profiles(d, broad_interest, enjoyment, self_efficacy, n_profiles = 3, model = 2)

plot_profiles(m3, to_center = TRUE)
```

To compare a range of profile solutions, use `compare_solutions()`

```{r}
compare_solutions(d, broad_interest, enjoyment, self_efficacy)
```

## Learn more

To learn more: 

* Browse the tidyLPA [website](https://jrosen48.github.io/tidyLPA/)

* *Read the Introduction to tidyLPA* [vignette](https://jrosen48.github.io/tidyLPA/articles/Introduction_to_tidyLPA.html), which has much more information on the models that can be specified with tidyLPA and on additional functionality

## Contact

As tidyLPA is at an early stage of its development, issues should be expected. If you have any questions or feedback, please do not hesitate to get in touch:

* By [email](mailto:jrosen@msu.edu)
* By [Twitter](http://twitter.com/jrosenberg6432)
* Through filing an issue on GitHub [here](https://github.com/jrosen48/tidyLPA)

Please note that this project is released with a [Contributor Code of Conduct](CONDUCT.md). By participating in this project you agree to abide by its terms.
