---
title: "Introduction to tidyLPA"
author: "Joshua M. Rosenberg"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Example

```{r, echo = F, message = F}
devtools::load_all(".")
```

```{r, eval = F}
library(tidyLPA)
```

Using the built-in `pisaUSA15` dataset (using just 200 observations for illustrative purposes) and variables for broad interest, enjoyment, and self-efficacy, we can quickly explore a number of models - both in terms of **the specification of the model** and **the number of profiles** using the  Bayesian Information Criteria (BIC) values.

```{r}
library(dplyr)
d <- pisaUSA15
d <- sample_n(pisaUSA15, 100)
compare_models_lpa(d, broad_interest, enjoyment, self_efficacy)
```

We can explore a three profile solution, with, for example, `4` profiles, and varying means across profiles, but equal variances and covariances (specified with model = `2`):

```{r, warning = F}
m3 <- create_profiles_lpa(d, broad_interest, enjoyment, self_efficacy, n_profiles = 4, model = 2)
plot_profiles_lpa(m3, to_center = TRUE)
```

See `?create_profiles_lpa` for a description for the models.

We could specify them as follows, whereby model `1` is for a model with varying means across profiles, but equal variances across profiles, and residual covariances fixed to zero, and model `3` is for a model with varying means, variances, and covariances across profiles:

```{r}
m3i <- create_profiles_lpa(d, broad_interest, enjoyment, self_efficacy, n_profiles = 3, model = 1)
```

This model has some trouble being fit, so is shown but is not run:

```{r, eval = F}
m3ii <- create_profiles_lpa(d, broad_interest, enjoyment, self_efficacy, n_profiles = 3, model = 3)
```

To test profile solutions for a specific model, you can carry out a bootstrapped likelihood-ratio test:

```{r}
bootstrap_lrt(d, broad_interest, enjoyment, self_efficacy)
```

For use in subsequent analyses, we can extract the posterior probabilities by setting `return_posterior_probs` to `TRUE`:

```{r}
m3 <- create_profiles_lpa(d, broad_interest, enjoyment, self_efficacy, n_profiles = 3, model = 2, return_posterior_probs = TRUE)
m3
```
